The Galactic Midplane Is Not a Plane: Implications for Dynamical Analysis with {\em Gaia} Data and Beyond

### 2019-04-26 paper/ms.tex
changes to conclusions wording

To further understand the effect of the midplane error, we also plot the
fractional error in $J_z$ as a function of $J_z$ for $z$ offsets of $10, 50,
100\,\pc$ (orange, cyan, and green, respectively) in
Figure~\ref{fig:dJz_fun_Jz}. For a thin disk-like orbit
($J_z\sim0.5\,\actunit$), even a $10\,\pc$ offset in $z$ is enough to induce a
$\sim25\%$ error in $J_z$. For larger values of $J_z$, the fractional errors
are suppressed, but the induced error can still be large depending on how
great the $z$ offset is. We also plot the epicylic prediction for $\Delta J_z
/ J_z$ from Equation~\ref{eq:actions_epi_error} as dashed lines for each $z$
offset. We find that the epicyclic approximation matches the numerical
estimate quite well.

@@

The second issue, which is related to the point of this paper, is that based
on the results in our previous sections computing $\bm{J}_i$ for a field star
and directly comparing it to  $\bm{J}_c$ will have significant systematic
error if the cluster and field star are far enough apart to have significantly
different local midplanes (see Figure \ref{fig:range_deltaphi}). Since our
criterion for cluster membership (distance in action space) scales with the
cluster mass, the region where the search will succeed varies with cluster
mass as well. To illustrate this effect, we assume a cluster mass $m_c$ and
estimate the expected action-space dispersion from
Equation~\ref{eq:action_disp}. We then consult
Figure~\ref{fig:many_orbit_wrong_ref} to determine the midplane offset
necessary in order for the induced error in the vertical action to be the same
as the expected action-space dispersion. We perform this analysis for the
thin-disk-like and thick-disk-like orbits considered in
Section~\ref{ssec:quant}, assuming that the halo-like orbit is unlikely for
members of known open clusters (see Table~\ref{tab:real_clusters}). The result
for the allowable midplane offset for these two orbits in shown in
Figure~\ref{fig:cluster_offset} as a function of cluster mass.

@@

\begin{itemize}
\item We demonstrated that an inaccuracy in the Galactocentric coordinate
system induces time evolution in the {\em observed} actions
(Figures~\ref{fig:cartoon}~and~\ref{fig:one_orbit_wrong_ref}). We interpret
this time evolution as an error in the computed action
(Figure~\ref{fig:many_orbit_wrong_ref}).

\item  We focused on the effect that an
inaccuracy in the midplane has on the vertical action $J_z$. We found that in
order for there to be a $10\%$ error in $J_z$ (as defined by one half the middle
$90\uth$ percentile) one needs a midplane error of $\sim6\,\pc$ and
$\sim50\,\pc$ for a typical thin and thick disk orbit, respectively. The
fractional error is significantly less for halo orbits.

\item We pointed out that the error distribution of the actions induced by a
coordinate system error is very non-Gaussian. The distribution is bimodal with
{\em neither of the modes peaking at 
     null}. 
As a result, error
propagation of systematic coordinate system errors is 
     complex
when considering actions.

\item We proposed that dynamical modelling across large regions of the disk is
susceptible to this type of error, since the assumption that our local
Galactic midplane is the global Galactic midplane is not true {\em a priori}.

\item To demonstrate the previous point, we measured the local galactic
midplane along the Solar circle in three different high-resolution, zoom-in
simulations of Milky Way mass galaxies from the FIRE collaboration. We found
that the midplane varies by $\sim185, 162, 84\,\pc$ (middle $90\%$) for the
three galaxies we considered (\mi{}, \mf{}, and \mm{}, respectively).

@@

@@

While in this work we have focused on errors in action computation, all of our
conclusions also extend to studies of stars that simply rely on orbit
integration, since the computation of actions and orbit integrations are
essentially equivalent.

{\bf Add sentences about distribution function modeling}

Our main point is that the local midplane varies between different points in
the Galaxy. We do
not propose here what a modeler should use for the ``true'' global midplane,
since it depends on the particular problem being studied. Even without a
high-precision definition of the global Galactic midplane, current
observations from {\em Gaia} should soon permit a measurement of the real
azimuthal dependence of the midplane location.

>>>>

To further understand the effect of the midplane error, we also plot the
fractional error in $J_z$ as a function of $J_z$ for $z$ offsets of $10, 50,
100\,\pc$ (orange, cyan, and green, respectively) in
Figure~\ref{fig:dJz_fun_Jz}. For a thin disk-like orbit
($J_z\sim0.5\,\actunit$), even a $10\,\pc$ offset in $z$ is enough to induce a
$\sim25\%$ error in $J_z$. For larger values of $J_z$, the fractional errors
are suppressed, but the induced error can still be large depending on how
great the $z$ offset is. We also plot the epicylic prediction for $\Delta J_z
/ J_z$ from Equation~\ref{eq:Ji_err_mosttime} as dashed lines for each $z$
offset. We find that the epicyclic approximation matches the numerical
estimate quite well.

@@

The second issue, which is related to the point of this paper, is that based
on the results in our previous sections computing $\bm{J}_i$ for a field star
and directly comparing it to  $\bm{J}_c$ will have significant systematic
error if the cluster and field star are far enough apart to have significantly
different local midplanes (see Figures \ref{fig:midplane} and \ref{fig:midplane_chervin}). Since our criterion for cluster membership (distance in action space) scales with the
cluster mass, the region where the search will succeed varies with cluster
mass as well. To illustrate this effect, we assume a cluster mass $m_c$ and
estimate the expected action-space dispersion from
Equation~\ref{eq:action_disp}. We then consult
Figure~\ref{fig:many_orbit_wrong_ref} to determine the midplane offset
necessary in order for the induced error in the vertical action to be the same
as the expected action-space dispersion. We perform this analysis for the
thin-disk-like and thick-disk-like orbits considered in
Section~\ref{ssec:quant}, assuming that the halo-like orbit is unlikely for
members of known open clusters (see Table~\ref{tab:real_clusters}). The result
for the allowable midplane offset for these two orbits in shown in
Figure~\ref{fig:cluster_offset} as a function of cluster mass.

@@

\begin{itemize}
\item Inaccuracy in the Galactocentric coordinate
system induces phase-dependence in the actions calculated from the observed positions and velocities of stars
(Figures~\ref{fig:cartoon}~and~\ref{fig:one_orbit_wrong_ref}). Since stars' instantaneous phase-space positions are measured without prior knowledge of their orbital phases, this results in systematic error in the computed actions
(Figure~\ref{fig:many_orbit_wrong_ref}).

\item Inaccuracy in the midplane location most severely affects computation of the vertical action $J_z$.  A midplane offset of $\sim6\,\pc$ ($\sim50\,\pc$) for a typical thin (thick) disk orbit results in a  $10\%$ error in $J_z$ (as defined by one half the middle
$90\uth$ percentile). The fractional error is significantly less for halo orbits.

\item The distribution of systematic errors in the actions induced by a
coordinate system offset is very non-Gaussian. The distribution is bimodal with
{\em neither mode at null}. 
As a result, error propagation of coordinate system offsets is complex when considering actions, and is likely to significantly deform the action-space distribution function.

\item Dynamical modelling across large regions of the disk, over which the midplane location varies by more than the limits discussed above, is
susceptible to this type of systematic error, since the assumption that our local
Galactic midplane is the global Galactic midplane is not true {\em a priori}.

\item To demonstrate the previous point, we measured the local galactic
midplane along the Solar circle in three different high-resolution, zoom-in
simulations of Milky Way mass galaxies from the FIRE collaboration. We found
that the midplane varies as a function of azimuth by $\sim185, 162, 84\,\pc$ (middle $90\%$) for the
three galaxies we considered (\mi{}, \mf{}, and \mm{}, respectively).

@@

\item While in this work we have focused on systematic errors in action computation, all of our conclusions also extend to studies of stars that simply rely on orbit
integration, since the computation of actions and orbit integrations are
mathematically equivalent.

@@

Our main point is that the local midplane varies between different points in
the Galaxy, and that this variation can lead to significant systematic errors in the computation of actions under the assumption of a global axisymmetric potential. We do
not propose here what a modeler should use for the ``true'' global midplane,
since it depends on the particular problem being studied. Even without a
high-precision definition of the global Galactic midplane, current
observations from {\em Gaia} should soon permit a measurement of the real
azimuthal dependence of the midplane location. For some applications, such as those using actions as labels to group stars on similar orbits, using such a measurement to shift stars to a consistent midplane height as a function of azimuth before using a global axisymmetric approximation to the potential may be sufficient, although this ignores the \emph{dynamical} implications of shifts in the midplane height (which result from fluctuations in the local density). However, for other applications, such as the study of action diffusion which motivated this paper in the first place, a more extensive perturbative approach is likely to be needed. We plan to explore the mitigation of these effects in future work.
